### 2012-10-29 Docs/Thesis/abstract.tex
Edits

Functional Reactive Programming (FRP) is a promising class of abstractions for encoding interactive and time-dependent programs in
functional languages which has proven difficult to implement efficiently. Signal-function FRP is a subclass of FRP which does not provide
direct access to the time-varying values by the programmer, but rather provides signal functions as first-class objects in the
program. FRP systems provide values defined at all points in time ({\em behaviors} or {\em signals}) and values defined
at countably many points in time ({\em events}).

All signal-function implementations of FRP to date have utilized demand-driven or ``pull-based''  evaluation for both events and signals,
producing output from the FRP system whenever the consumer of the output is ready. This greatly simplifies the implementation of signal-function
FRP systems, but leads to inefficient and wasteful evaluation of the FRP system.

@@

I describe the implementation of a push-based signal-function FRP system. The semantics of the system are discussed, and its performance
and expressiveness for practical examples of interactive programs are compared to existing signal-function FRP systems through the
implementation of a networking application.

>>>>

Functional Reactive Programming (FRP)is a promising class of abstractions for
interactive programs. FRP systems provide values defined at all points in time
({\em behaviors} or {\em signals}) and values defined at countably many points
in time ({\em events}) as abstractions.

Signal-function FRP is a subclass of FRP which does not provide direct access to
the time-varying values to the programmer, but rather provides signal functions,
which are reactive transformers of signals and events, as first-class objects in the program.

All signal-function implementations of FRP to date have utilized demand-driven
or ``pull-based''  evaluation for both events and signals, producing output from
the FRP system whenever the consumer of the output is ready. This greatly
simplifies the implementation of signal-function FRP systems, but leads to
inefficient and wasteful evaluation of the FRP system when this strategy is
employed to evaluate events.

@@

I describe the implementation of a signal-function FRP system which applies
pull-based evaluation to signals and push-based evaluation to events (a
"push-pull" system). The semantics of the system are discussed, and its
performance and expressiveness for practical examples of interactive programs
are compared to existing signal-function FRP systems through the implementation
of a networking application.

### 2012-10-29 Docs/Thesis/background.tex
Fix figure captions
Edits

@@

The goal of FRP is to provide compositional and high-level abstractions for creating reactive programs. The key
abstractions of FRP are behaviors or signals\footnote{Behaviors are generally referred to as {\em signals} in signal function literature. This is unfortunate, since a signal
function may operate on events, signals, or some combination of the two.}, which are time-dependent values defined at every point in continuous time, and events, which are 
values defined at countably many points in time. An FRP system will provide functions to manipulate events and signals and to react
to events by replacing a portion of the running program in response to an event. Behaviors and events, or some abstraction
based on them, will be first class, in keeping with the spirit of functional programming. Programs implemented in FRP languages
should of course be efficiently executable. This has proven to be the main challenge in implementing FRP.

@@

The earliest and still standard formulation of FRP provides two primitive type
constructors: {\tt Behavior} and {\tt Event}, together with combinators that
produce values of these types. The easiest semantic definition for these types
is given in Figure~\ref{figure:classic_frp_semantic_types}\footnote{The list constructor {\tt []} should be considered to contain infinite as well as finite lists.}.

\begin{figure}
\begin{code}
type Event a = [(Time, a)]
type Behavior a = Time -> a
\end{code}
\hrule
\caption{Semantic types for Classic FRP}
\label{figure:classic_frp_semantic_types}
\end{figure}

@@

switcher :: Behavior a -> Event (Behavior a) -> Behavior a
switcher b [] t = b t
switcher b [(to, bo)] t = if t < to
                          then b t
                          else bo t
\end{code}
\hrule
\caption{An obvious, yet inefficient and problematic, implementation of
Classic FRP}
\label{figure:classic_frp_simple_implementation}
\end{figure}

Obvious problems include the lack of ordering enforcement on event occurrences,
the necessity of waiting for the first event occurrence in a switch, and the
necessity of maintaining the full history of a behavior for evaluation.

@@

The first implementation of FRP outside the Haskell language was Frapp\'{e}, an
FRP implementation for the Java Beans framework. Frapp\'{e} built on the notion
of events and ``bound properties'' in the Beans framework, providing abstract
interfaces for FRP events and behaviors, and combinators as concrete classes
implementing these interfaces. The evaluation of Frapp\'{e} used Bean components
as sources and sinks, and the implementation of Bean events and bound properties
to propagate changes to the network~\cite{Courtney2001-2}.

>>>>

In contrast, the more popular imperative model of programming takes statements,
or actions which modify the state of the world, as the primary building blocks
of programs. Such programs have inherently sequential control flow, and require
reasoning about side-effects. They are thus inherently resistant to
compositional reasoning.

@@

The goal of FRP is to provide compositional and high-level abstractions for creating reactive programs. The key
abstractions of FRP are behaviors or signals\footnote{Behaviors are generally referred to as {\em signals} in signal function literature. This is unfortunate, since a signal
function may operate on events, signals, or some combination of the two.}, which are time-dependent values defined at every point in continuous time, and events, which are 
values defined at countably many points in time. An FRP system will provide combinators to manipulate events and signals and to react
to events by replacing a portion of the running program in response to an event. Behaviors and events, or some abstraction
based on them, will be first class, in keeping with the spirit of functional programming. Programs implemented in FRP languages
should of course be efficiently executable, but this has proven to be the main challenge in implementing FRP.

@@

The earliest and still standard formulation of FRP provides two primitive type
constructors: {\tt Behavior} and {\tt Event}, together with combinators that
produce values of these types. The easiest semantic definition for these types
is given in Figure~\ref{figure:classic_frp_semantic_types}\footnote{The list type constructor {\tt []} should be considered to contain infinite as well as finite lists.}.

\begin{figure}
\begin{code}
type Event a = [(Time, a)]
type Behavior a = Time -> a
\end{code}
\hrule
\caption{Semantic types for Classic FRP.}
\label{figure:classic_frp_semantic_types}
\end{figure}

@@

switcher :: Behavior a -> Event (Behavior a) -> Behavior a
switcher b [] t = b t
switcher b [(to, bo)] t = if t < to
                          then b t
                          else bo t
\end{code}
\hrule
\caption{An obvious, yet inefficient and problematic, implementation of
Classic FRP.}
\label{figure:classic_frp_simple_implementation}
\end{figure}

There are several obvious problems with this implementation of course, but it
suffices to show the intuition behind the classic FRP model. Problems in this
implementation which are addressed by real implementations include the necessity
of waiting for event occurrences, the necessity of maintaining the full history
of a behavior in memory, and the lack of an enforced order for event
occurrences.

@@

The first implementation of FRP outside the Haskell language was Frapp\'{e}, an
FRP implementation for the Java Beans framework. Frapp\'{e} built on the notion
of events\footnote{These are semantic objects to which callbacks may be added,
and not events in the FRP sense, though they are related.} and ``bound
properties'' in the Beans framework, providing abstract interfaces for FRP
events and behaviors, and combinators as concrete classes implementing these
interfaces. The evaluation of Frapp\'{e} used Bean components as sources and
sinks, and the implementation of Bean events and bound properties to propagate
changes to the network~\cite{Courtney2001-2}.

### 2012-10-29 Docs/Thesis/introduction.tex
Edits

Most of the useful programs which programmers are asked to write must react to inputs which are not available at the time the program starts,
and produce effects at many different times throughout the execution of the program. Examples of these programs include GUI applications,
web browsers, web applications, video games, multimedia authoring and playback tools, operating system shells and kernels, servers,
robotic control programs, and many others. This ability is called reactivity.

@@

The first challenge was addressed by representing FRP programs, not as compositions of signals and events, but as compositions of
transformers of behaviors and events called {\em signal functions}. By not permitting direct access to behaviors or events, but
representing them implicitly, signal functions prohibit accessing past or future time values, only allowing access to values in the closure of the signal function
and current input values. Signal function FRP programs are written by composing signal functions, and are evaluated using a function which provides input to the
signal function and acts on the output values. This evaluation function is elided in existing literature on signal functions. 
A further advantage of signal function FRP programs is that they are more compositional, since additional signal functions may be composed
on the input or output side of a signal function, rather than simply the output side in classic (non-signal-function) FRP systems.

The second challenge was addressed for classic FRP by the creation of a hybrid push-pull FRP system~\cite{Elliott2009}. This system relied
on the distinction between reactivity and time dependence to decompose behaviors into phases of constant or simply time-dependent values,
with new phases initiated by event occurrences. However, this implementation did not address concerns of causality or compositionality.
Further, the existence of events as first-class values in classic FRP forced the use of an impure and resource-leaking technique to compare
events to determine which had the first occurrence after the current time. This technique made use of the Haskell {\tt unsafePerformIO} function
to fork two threads to compare the improving values of the next occurrence time for two events being merged. Further, since the classic FRP
interface permits access to only the output of a behavior or event, and is always bound to its implicit inputs, the best the push-based implementation
could do is to block when no computation needs to be performed. The computation cannot be suspended entirely as a value representation.

To address both challenges in a FRP implementation, it seems desirable to combine the approaches, creating a push-based
signal-function FRP system. The current implementation approach for signal-function FRP is inherently pull-based~\cite{Nilsson2002}, but
recent work on N-ary FRP~\cite{Sculthorpe2011}, a variant of signal-function FRP which enumerates inputs and distinguishes between events and behaviors, provides a representation which suggests
a method of push-based evaluation. In previous implementations of FRP, events were simply
option-valued signals. This approach has the advantage of being simple to implement in a pull-based setting, but does not have an obvious
semantics for event merging, and necessitates pull-based evaluation, since there is no way to separate the events which may be pushed from
the truly continuous values which must be repeatedly sampled.

This thesis describes the design and implementation of Time-Flies, a push-pull signal-function FRP system. This system is presented as a library
of combinators for constructing and evaluating FRP programs.

>>>>

Most of the useful programs which programmers are asked to write must react to inputs which are not available at the time the program starts,
and produce effects at many different times throughout the execution of the program. Examples of these programs include GUI applications,
web browsers, web applications, video games, multimedia authoring and playback tools, operating system shells and kernels, servers,
robotic control programs, and many others. This attribute of a program is called
reactivity.

@@

The first challenge was addressed by representing FRP programs, not as
compositions of signals and events, but as compositions of transformers of
behaviors and events called {\em signal functions}. By not permitting direct
access to behaviors or events, but representing them implicitly, signal
functions prohibit accessing past or future time values, only allowing access to
values in the closure of the signal function and current input values. Signal
function FRP programs are written by composing signal functions, and are
evaluated using a function which provides input to the signal function and acts
on the output values. This evaluation function is elided in existing literature
on signal functions, but we provide a documented and improved interface for
evaluating signal functions. A further advantage of signal function FRP programs
is that they are more composable, since additional signal functions may be
composed on the input or output side of a signal function, rather than simply
the output side in classic (non-signal-function) FRP systems.

The second challenge was addressed for classic FRP by the creation of a hybrid
push-pull FRP system~\cite{Elliott2009}. This system relied on the distinction
between reactivity and time dependence to decompose behaviors into phases of
constant or simply time-dependent values, with new phases initiated by event
occurrences. However, this implementation did not address concerns of causality
or compositionality. Further, the existence of events as first-class values in
classic FRP forced the use of an impure and resource-leaking technique to
compare events when merging, to determine which had the first occurrence after
the current time. This technique made use of the Haskell {\tt unsafePerformIO}
function to fork two threads to compare the improving values of the next
occurrence time for two events being merged. Further, since the classic FRP
interface permits access to only the output of a behavior or event, and is
always bound to its implicit inputs, the best the push-based implementation
could do is to block when no computation needs to be performed. The computation
cannot be suspended entirely as a value representation.

To address both challenges in a FRP implementation, we to combine the approaches
and demonstrate a push-based signal-function FRP system. The current
implementation approach for signal-function FRP is inherently
pull-based~\cite{Nilsson2002}, but recent work on N-ary
FRP~\cite{Sculthorpe2011}, a variant of signal-function FRP which enumerates
inputs and distinguishes between events and behaviors, provides a representation
which suggests a method of push-based evaluation. In previous implementations of
FRP, events were simply option-valued signals. This approach has the advantage
of being simple to implement in a pull-based setting. However, it does not have
an obvious semantics for event merging. Further, it necessitates pull-based
evaluation, since there is no way to separate the events which may be pushed
from the truly continuous values which must be repeatedly sampled.

This thesis describes the design and implementation of TimeFlies, a push-pull signal-function FRP system. This system is presented as a library
of combinators for constructing and evaluating FRP programs.
